---
title: "Dates and Times"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r global_options, include=FALSE}
# The following code is a basic setup of options for your document
knitr::opts_chunk$set(echo = TRUE,
                      eval = TRUE, 
                      warning = TRUE,
                      message = FALSE,
                      fig.align = "center", 
                      R.options = list(max.print=50))

# Edit the file starting below

```

In this worksheet, we will:

-   Parse dates and times from strings into R-compatible formats using `lubridate`
-   Manipulate dates and times, including extracting specific components (year, month, day, hour, etc.)
-   Calculate differences between two dates/times
-   Visualize time-based data trends and summarize data over time units

## 0. Datasets and Libraries

Let's load `tidyverse` which contains the `lubridate` package and `nycflights13` for datasets:

```{r, warning=FALSE}
# Load packages
library(tidyverse)
library(nycflights13)
```

Here are the main datasets we will manipulate today:

```{r}
# Take a look at txhousing and flights
head(txhousing)
head(flights)
```

In what format are dates/times reported in each dataset?

**Write sentences here.**  
Only time_hour is in a date/time format

## 1. Parsing dates and times

Parsing dates and times means converting a string or numeric representation of a date or time into a proper object that R can understand and work with.

### a. R formats

Here are two typical date/time formats in R:

```{r}
# Look at what date/time is today/now!
today()
now()
```

```{r}
# How R identifies these formats
class(today())
class(now())
```

By default, R considers:

-   *dates* as "yyyy-mm-dd" (year-month-day)

-   *times* as "hh:mm:ss" (hours:minutes:seconds)

-   *date/times* as "yyyy-mm-dd hh:mm:ss"

Here is an example of a date entered in the year-month-day but does R recognize it as a date?

```{r}
# Check the format of the date entered
class("2024-10-03")
```

We can convert a character/string and convert it as a date in an R format:

```{r}
# year, month, day
ymd("2024-10-03")

# Check the format now
class(ymd("2024-10-03"))
```

Here are some examples of other formats of dates than can be converted:

```{r}
# day, month, year
dmy("3.10.2024")
dmy("3/10/2024")
dmy("3-10-2024")
dmy("the 3rd of October 2024")
dmy("03-octobre-2024") # this one did not work, why?

# month, day, year
mdy("10/3/2024")
mdy("October 3rd, 2024")
```

Similarly, we can convert strings into time:

```{r}
# date in year, month, day and time
ymd_hms("2024-10-03 03:15:00 PM")

# also check other date functions with _hms or _hm, or simply the function hm() and hms()
hm("03:15 PM")
hms("03:15:00 PM")
```

### b. Combining date/time components

We can combine the different parts of a date with `make_date()` or also add time with `make_datetime()`.

```{r}
# Combine year and month into a date
txhousing |>
  mutate(new_date = make_date(year, month))
```

By default, the day on the date was set to the first day of the month.

### c. Extracting part(s) of the date

On the contrary, we might want to extract some specific date/time information from a date:

```{r}
# Extract year, month, day and time
year(now())
month(now())
week(now())
day(now())
wday(now()) # what is that? - Weekday (1 = Sunday, 2 = Monday, ...)

hour(now())
minute(now())
second(now())
```

Check the `label` and `abbr` options for `month()` and `wkday()`:

```{r}
# Convenient options
month(now(), label = TRUE, abbr = FALSE)
wday(now(), , label = TRUE, abbr = TRUE)
```

-------------------------------------------------------------------------------

#### **Try it! In the `flights` dataset, extract the information of the weekday from the `time_hour` variable. Does each day have the same amount of flights?**

```{r}
# Write and submit code here!
flights |>
  mutate(weekday = wday(time_hour, , label = TRUE, abbr = TRUE)) |>
  ggplot() + geom_bar(aes(x = weekday))
```

**Write sentences here.**

-------------------------------------------------------------------------------

## 2. Manipulating dates and times

### a. Finding differences between dates and times

We can find date/time differences with `difftime()`:

```{r}
# How many days between now and the first day of the year?
difftime(now(), mdy_hms("1-1-2024 00:00:00 am"), units = "days")

# What if we want to find the difference with another unit?
difftime(now(), mdy_hms("1-1-2024 00:00:00 am"), units = "weeks")
difftime(now(), mdy_hms("1-1-2024 00:00:00 am"), units = "hours")
difftime(now(), mdy_hms("1-1-2024 00:00:00 am"), units = "mins")
difftime(now(), mdy_hms("1-1-2024 00:00:00 am"), units = "secs")
```

Note that the output reports the time difference with a unit. If we would like to find the value of the difference, we can use the function `as.numeric()`:

```{r}
# Report only a value
as.numeric(difftime(now(), mdy_hms("1-1-2024 00:00:00 am"), units = "days"))
```

### b. Summarizing date/time data

Depending on the level of detail we would like to focus on, we can aggregate the data by specific time units. For example, we can compare summaries over years, months, days of the week, or by the hour, minute, second.

-------------------------------------------------------------------------------

#### **Try it! We looked at the number of `flights` per day before. Compare the number of flights at another time unit. Do you notice any differences?**

```{r}
# Write and submit code here!
flights |>
  ggplot() + geom_bar(aes(x = minute)) +
  scale_x_continuous(breaks = seq(0, 60, by = 10))
```

**Write sentences here.**

-------------------------------------------------------------------------------

We can also represent the values of a variable over time:

```{r}
# Comparing sales over time
txhousing |>
  group_by(date) |>
  summarize(total_sales = sum(sales, na.rm = TRUE)) |>
  ggplot() + geom_line(aes(x = date, y = total_sales))
```

And compare if there is the same pattern over a repeating time unit (for example, months repeat every year):

```{r}
# Comparing monthly sales for each year
txhousing |>
  group_by(year,month) |>
  summarize(total_sales = sum(sales, na.rm = TRUE)) |>
  ggplot() + geom_line(aes(x = month, y = total_sales)) +
  facet_wrap(~year, nrow = 1)
```

-------------------------------------------------------------------------------

#### **Try it! Compare the maximum distance for a flight per hour of the day. When do longer flights depart from NYC airports?**

```{r}
# Write and submit code here!
flights |>
  mutate(weekday = wday(time_hour, , label = TRUE, abbr = TRUE)) |>
  group_by(hour) |>
  slice_max(distance, n = 1) |>
  ggplot() + geom_point(aes(x = hour, y = distance)) +
  scale_x_continuous(breaks = seq(0, 24, by = 1))
```

**Write sentences here.**

-------------------------------------------------------------------------------

### c. A few remarks

Here are some common pitfalls to look out for:

-   Different date formats (e.g., MM/DD/YYYY vs. DD/MM/YYYY) can lead to incorrect parsing. Always specify the date format explicitly when converting strings to dates.

-   Take into account that not all years are 365 days (leap years), not all days are 24 hours (daylight saving time), and not all months have the same amount of days. Most `lubridate` functions are designed to take those facts into account when manipulating dates/times.

-   The time is not the same depending on where the data was collected. Convert dates/times between time zones with some `lubridate` functions such as `with_tz()`.

```{r}
# Time in France
now(tzone = "Europe/Paris")
# Convert to our current time zone
with_tz(now(tzone = "Europe/Paris"), tzone = "America/Chicago")
```

-------------------------------------------------------------------------------

## Group Practice

Let's explore birthdays from students in this class!

1.  Enter your birthday (including year) on this [spreadsheet](https://docs.google.com/spreadsheets/d/1DDB1l_IyVz7yVPKJKt9bQd4pCV3gJnIwU41a19zRd94/edit?usp=sharing). Once everyone has entered their birthdays, download the spreadsheet as a `.csv` file and import it in `R`.

2.  Does any birthday need to be recoded? Parse the birthday into a date format that R can work with.

3.  Make an analysis of the birthdays and paste any plot or statistics you create [here](https://docs.google.com/presentation/d/1cDttg4oytMW-S_dhXDSuymQWPsqzzYo1ix9pI8sNH18/edit?usp=sharing).

```{r echo = FALSE}
# Group practice
bday <- read.csv("Birthdays - Sheet1.csv")
bday |>
  mutate(new_bdays = mdy(birthdays)) |>
  mutate(year = year(new_bdays), month = month(new_bdays), day = day(new_bdays)) |>
  group_by(month) |>
  ggplot() + geom_bar(aes(x = year, fill = as.factor(day))) + 
  theme(legend.direction = "horizontal") + 
  labs(fill = "Day", x = "Year", y = "Number of Birthdays", title = "Number of Birthdays per year")
bday |>
  mutate(new_bdays = mdy(birthdays)) |>
  mutate(year = year(new_bdays), month = month(new_bdays), day = day(new_bdays)) |>
  group_by(month) |>
  ggplot() + geom_point(aes(x = month, y = day)) +
  scale_x_continuous(breaks = seq(0, 12, by = 1))
bday |>
  mutate(new_bdays = mdy(birthdays)) |>
  mutate(year = year(new_bdays), month = month(new_bdays), day = day(new_bdays)) |>
  group_by(new_bdays) |>
  summarize(n = n()) |>
  filter(n > 1)
```
