---
title: "More Data Wrangling"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
---

In this worksheet, we will:

-   Recode some variables for easier visualizations
-   Discuss how to handle missing values

## 0. Dataset and Library

We will again consider the built-in dataset `txhousing`, available in the `tidyverse` package:

```{r}
# Upload the package
library(tidyverse)

# Take a look at the first few rows of the dataset
head(txhousing) 
```

Let's further manipulate this dataset and create visualizations.

## 1. Recoding variables

There are many reasons why we may want to recode some variables. This list is not exhaustive!

### a. Log-transformation

When a variable is highly skewed, it might be difficult to "see" the variation:

```{r}
# Check this boxplot
txhousing |>
  ggplot() +
  geom_boxplot(aes(x = sales))
```

For distributions of numeric variables that are heavily right-skewed like this one we can apply a log-transformation:

```{r}
txhousing |>
  # Transform the sales variable
  mutate(log_sales = log(sales)) |>
  # Use this new variable in a ggplot
  ggplot() +
  geom_boxplot(aes(x = log_sales))
```

What is the unit of this new variable?

**Write sentences here.**

Note that this log-transformation only works for right-skewed data and can only be applied to values greater than 0.

### b. Numeric variable considered as a categorical variable

Some variables are coded numerically but would be better to represent as a category:

```{r}
# Check this plot
txhousing |>
  ggplot(aes(x = year, y = sales)) +
  geom_smooth(se = FALSE)

txhousing |>
  # Color by year as a numeric variable
  ggplot(aes(x = year, y = sales, color = month)) +
  geom_smooth(se = FALSE)

txhousing |>
  # Consider year as a factor (a type of categorical variable)
  mutate(month_fct = as.factor(month)) |>
  # Color by year as a categorical variable
  ggplot(aes(x = year, y = sales, color = month_fct)) +
  geom_smooth(se = FALSE)
```

The `as.numeric()` would do the opposite: make a variable numeric.

### c. Recoding values based on conditions

We can recode some values based on conditions with `case_when()`:

```{r}
txhousing |>
  # Recode months into 4 quarters
  mutate(month_cat = case_when(
    month <= 3 ~ "1st quarter",
    4 <= month & month <= 6 ~ "2nd quarter",
    7 <= month & month <= 9 ~ "3rd quarter",
    10 <= month & month <= 12 ~ "4th quarter")) |>
  # Use this new variable in a ggplot
  ggplot(aes(x = year, y = sales, color = month_cat)) +
  geom_smooth(se = FALSE)
```

Or we can recode each value with `recode()`:

```{r}
txhousing |>
  # Recode months with Spanish names
  mutate(month_es = recode(month,
                           `1` = "enero", `2` = "febrero", `3` = "marzo",
                           `4` = "abril", `5` = "mayo", `6` = "junio",
                           `7` = "julio", `8` = "agosto", `9` = "septiembre",
                           `10` = "octubre", `11` = "noviembre", `12` = "diciembre")) |>
  # You can control in which order the levels of a factor variable would appear
  mutate(month_es = factor(month_es, 
                           levels = c("enero", "febrero", "marzo", 
                                      "abril", "mayo", "junio",
                                      "julio", "agosto", "septiembre", 
                                      "octubre", "noviembre", "diciembre"))) |>
  # Use this new variable in a ggplot
  ggplot(aes(x = month_es, y = sales)) +
  geom_bar(stat = "summary", fun = sum)
```

## 2. Handling missing values

We talked about removing missing values in calculations with `na.rm = TRUE` for example. But simply ignoring missing data without trying to understand why the data might be missing could introduce some bias in our visualizations and statistics.

### a. Missing values never go unnoticed

When calculating statistics:

```{r}
txhousing |>
  # Find mean and correlation
  summarize(mean_sales = mean(sales),
            correlation = cor(sales,listings))
```

When making visualizations:

```{r}
# Check the Warning message
txhousing |>
  ggplot() +
  geom_boxplot(aes(x = sales))
```

Before omitting the missing values, we should check how many values are missing and if there is any pattern in the missing data:

```{r}
txhousing |> 
  # Count missing values for each variable
  summarize_all(~ sum(is.na(.)))
```

```{r}
txhousing |>
  # Split by year/month
  group_by(year,month) |>
  # Count missing values for each variable
  summarize_all(~ sum(is.na(.)))
```

Note that it looks like more recent data is not missing as much as older data.

```{r}
txhousing |>
  # Split by city
  group_by(city) |>
  # Count missing values for each variable
  summarize_all(~ sum(is.na(.)))
```

Some cities don't have any data missing, others have only a few, and others have a lot!

### b. Values that were omitted

If we have monthly data in `txhousing` for 46 cities, from 2000 to 2015, how many rows should we expect?

```{r}

```

The missing rows are due to the data only being reported until July of 2015. It would not make sense to compare yearly data in 2015 to other years.
